An Empirical Analysis of the Internet Engineering Task Force with Computational Methods

### 2024-10-07 paper.tex
Tweaks

These two stages -- matching with the Datatracker, and merging previously
seen names and addresses -- accounts for the majority (60\%) of messages.
If an email's sender is not in the Datatracker, and the name and address
has not been previously seen, a new person identifier is generated. This
accounts for a small portion (~10\%) of messages. This is reasonable, given
that Datatracker profiles have become necessary for many of the IETF's
day-to-day activities.

@@

%..................................................................................................
\pb{Reproducibility and data access:}
We have developed a library, \texttt{ietfdata}, that fetches the RFC
Index, communicates with the IETF Datatracker API, and retrieves messages
from the IETF IMAP mail archives. This library appropriately regulates
access, caches data to minimise the impact on the infrastructure, and
performs necessary post-processing. This library is available as open
source and Ongoing development is coordinated on GitHub.%
\footnote{\url{https://github.com/glasgow-ipl/ietfdata}}

@@

The data we analyse is extracted from public IETF archives and APIs.  We
have taken steps to ensure ethical compliance.  To ensure that our access
to these services does not cause operational problems for the IETF, we are
in regular contact with the IETF Tools Team and Secretariat, as well as the
operators of the Datatracker and mailing list archive.  We have
discussed our work with IETF leadership (IETF and IAB Chairs, the IETF
Executive Director, and the IRTF Chair, who is a co-author on this paper)
and the RFC Editor to ensure that our access falls within their acceptable
use policies.

@@

\begin{figure}
  \centering
  \includegraphics[width=\figureWidthOneColumn]{figures/drafts-by-date.pdf}
  \caption{
    Number of active drafts over time. Internet-Drafts, working documents
    of the IETF for work that was intended for eventual publication as an
    RFC, were introduction at the 13th IETF meeting in 1989.
  }
  \label{fig:drafts-by-date}
\end{figure}

@@

We observe that the number of active drafts correlates more strongly with
the number of active working groups (Figure \ref{fig:pub_wgs_yearly}) than
with RFC publication (Figure \ref{fig:rfcs_by_area}).  This is largely to
be expected: an Internet-Draft is a work-in-progress document that is
proposed for adoption by, and discussion in, an IETF working group prior to
eventual publication as an RFC.  Competing proposals are often submitted as
drafts, to allow a working group to review them and make a decision how to
proceed, and hence not all drafts become RFCs. As we discuss in
\S\ref{sec:success-factors:documents}, draft adoption is a significant
milestone.

@@

Figure~\ref{fig:keyword_usage_rates} further confirms the growing complexity
of RFCs, showing how the use of normative language keywords has evolved
over time.  Keywords are used in RFCs to indicate the normative requirements
an RFC imposes on implementations. Figure~\ref{fig:keyword_usage_rates}
shows the total number of occurrences of each of the RFC 2119~\cite{RFC2119}
keywords (i.e., MUST, MUST NOT, REQUIRED, SHALL, SHALL NOT, SHOULD, SHOULD
NOT, RECOMMENDED, MAY, OPTIONAL), divided by the page count of the RFC. As
shown, the median number of keywords per page grew from 2001 through to
2010, indicating a growing number of requirements being expressed in RFCs,
before plateauing in recent years.

@@

These shifts in affiliation reflect changes in fortune and commercial
success of the companies. They also show the ongoing importance of IETF
standards, with new companies choosing to send participants: while we do
not demonstrate any causal link, it is encouraging that commercially
successful companies opt to enable their employees to actively participate
in the IETF. Care is needed to ensure that this relevance is maintained,
however: we observe that the author pool has grown less diverse in terms of
companies that are represented. 35.4\% of authors came from the top 10
affiliations in the dataset in 2020, compared with 25.6\% in 2001.

@@

As we showed in Section \ref{sec:org-dyn:participants}, protocol
standardisation is an inherently social process with most day-to-day work
happening on public mailing lists.  We are specifically
interested in better understanding how influence is distributed across
stakeholders and how it might affect the standardisation process.  This is
of critical societal importance: the IETF has a major impact on global
Internet technologies, and understanding the social processes involved
would give us insight into not only the driving forces behind
standardisation, but also its resilience to the loss of major participants.
In this section we consider the following research questions: 
\begin{enumerate}
  \item How centralised is the active IETF community, and to what extent
    is it reliant on a small core of participants? 
  \item How do the most influential participants behave? 
  \item How does influence (determined by mailing list participation)
    relate to wider impacts throughout the IETF?
  \item Does the organisational affiliation of participants also
    influence the innovation (adoption of new work) within IETF?
\end{enumerate}

@@

We have identified that influential authors write more drafts
(\S\ref{subsec:impact_influence}), but are these drafts more likely to
see success?  That is, does the presence of a well-connected, influential,
author make it more likely that an Internet-draft is adopted by a working
group for review and development\footnote{There are occasional document,
known as Area Director sponsored drafts, that are published as standards
track RFCs without going through the working group process, but these too
rare ($\sim$3\% of RFCs) to significantly affect our analysis.} and
eventually published as a standards track RFC.  Using information from the
IETF Datatracker we compile a dataset for the \emph{adoption} of a draft by
a working group, comprising 11632 drafts, and employ logistic regression to
analyse factors that affect working group adoption of draft.

@@

  \item We remove collinearity by using the Variance Inflation Criterion
    (VIF), removing all features with a VIF value above 5.

@@

We further see that some of the topics extracted by LDA are also useful.
For example, Topic 13 is characterised by a cluster of terms associated
with MPLS, a widely deployed routing protocol. We find a negative
correlation with this topic, but this is likely because there is a
significant number of RFCs that propose modifications or additional
features for the protocol, some of which do not see deployment. Overall,
most of the results are in line with expectations, but more annotated data
is required for further insights.

@@

\pb{Documents, Demographics, Communication Patterns:}
To the best of our knowledge, were the first to perform a comprehensive
statistical analysis of IETF activities considering trends in document
publication (\S\ref{sec:trends-documents}), author demographics
(\S\ref{sec:trends-demographics}), and communication patterns.
\S\ref{sec:org-dyn:participants}. That said, there have been several
prior efforts that have focused on specific aspects of the IETF:
Jari Arkko, a former Chair of the IETF, maintains a website
(\url{https://www.arkko.com/tools/stats.html}) that provides various
statistics about the IETF, including about its documents, authors and their
affiliations. The Datatracker also provides some statistics on documents
and authors (\url{https://datatracker.ietf.org/stats/}).  Attendance
statistics are reported in the IETF plenaries (available in the proceedings
at \url{https://www.ietf.org/meeting/past/}).

@@

\pb{Influence and Impact:}
Our work to characterise participant influence and impact (\S\ref{sec:org-dyn:influence})
builds on a rich literature regarding influence in online social networks
and other communities.
Studies of Twitter~\cite{cha2010measuring,weitzel2012measuring,anger2011measuring},
Instagram~\cite{zarei2020characterising} and decentralised social networks
\cite{harissigmetrics, hassanpleroma}. For example, \cite{ye2010measuring}
measured the impact of follower influence on message propagation. Others
have focused on devising metrics to capture influence: similarly
to~\cite{kourtellis2013identifying,weitzel2012measuring,sole2014centrality},
we use betweenness centrality as a metric of influence.

@@

\pb{Temporal Patterns in Communication:}
Analysing the temporal patterns in communication, as we do for the IETF
organisational hierarchy in \S\ref{sec:org-dyn:hierarchy}, is an emerging
field of research~\cite{masuda2016guide}:

Some have analysed the time evolution of important nodes in a
network~\cite{fire2020rise}, whether their neighbours are similarly
important~\cite{pedreschi2022temporal} and some others the correlations of
the trajectory of snapshots of networks over
time~\cite{lacasa2022correlations}. In this paper we bring new analysis
techniques, that expand on the idea of tracking the evolution of node
importance, to an email list corpus.

@@

The effects of hierarchy on the ``voice" of individuals within an
organisation is a key aspect~\cite{pfrombeck2023hierarchy} of
Organisational Behaviour (OB) research. A steep~\cite{hussain2019voice}
hierarchy can have a negative effect on voice by reducing the variety of
perspectives~\cite{anderson2010functions} in higher levels, or positive
effect where decisions can be made quickly. However, a diffused hierarchy
may also have negative effects~\cite{hussain2019voice} on voice by
contributing to lower level ``bystanders" who let the higher levels do the
talking, or a positive effect with a variety of approaches to management.
Similarly, a large ``power distance" can cause problems to lower level
voice if the higher levels are
``authoritarian"~\cite{duan2018authoritarian}, however if they are
``benevolent"~\cite{li2021does,guo2020inclusive} then power distance is
less of a problem. We aim to test these ideas in regard to the voluntary
and collaborative nature of the IETF organisation, whereas the literature
mostly focuses on traditional organisations.

@@

%..................................................................................................
\pb{Success Factors:}
The work in \S\ref{sec:success-factors} builds, in part, on Nikkhah et
al.~\cite{nikkhah2017statistical} by statistically exploring RFC adoption.
We build on this work with improved modelling, and by incorporating
additional features that characterise not only the RFCs themselves but also
the standardisation process that produced them.

@@

We explored the key standardisation activities within the IETF, considering
trends in document production, complexity, and correctness
(\S\ref{sec:trends-documents}); the people involved in Internet standards
development and how the demographics of the community have shifted over
time (\S\ref{sec:trends-demographics}); interaction between participants,
who has influence and impact in the IETF and how that influence is evident
in their activities, and the dynamics of communications within the IETF
(\S\ref{sec:org-dyn}); use of language, to explore whether there are
differences in linguistic patterns and language use in more influential
participants, and to consider how the use of language reflects the
consensus-driven nature of the standards development process
(\S\ref{sec:language}); and factors that lead to success, exploring what
makes a document likely to be published as an RFC and what makes a
successful participant (\S\ref{sec:success-factors}).

@@

  \item While authorship is becoming more diverse as the IETF grows, there
    are still significant issues with representation from certain regions,
    including Africa and South America. If the IETF is to become more
    representative, then further effort is required to build more diverse
    engagement, both in terms of geography, organisational affiliation, and
    gender diversity \cite{kaeo:2023:women}.

>>>>

These two stages -- matching with the Datatracker, and merging previously
seen names and addresses -- accounts for the majority (60\%) of messages.
If an email's sender is not in the Datatracker, and the name and address
has not been previously seen, a new person identifier is generated. This
accounts for a small portion (10\%) of messages. This is reasonable, since
Datatracker profiles are necessary for many of the IETF's day-to-day
activities.

@@

%..................................................................................................
\pb{Reproducibility and data access:}
We have developed a library, \texttt{ietfdata}, that fetches the RFC
Index, communicates with the IETF Datatracker API, and retrieves messages
from the IETF IMAP mail archives. This library appropriately regulates
access, caches data to minimise the impact on the infrastructure, and
performs necessary post-processing. This library is available as open
source and ongoing development is coordinated on GitHub.%
\footnote{\url{https://github.com/glasgow-ipl/ietfdata}}

@@

The data we analyse is extracted from public IETF archives and APIs.  We
have taken steps to ensure ethical compliance.  To ensure that our access
to these services does not cause operational problems for the IETF, we are
in regular contact with the IETF Tools Team and Secretariat, as well as the
operators of the Datatracker and mailing list archive.  We have
discussed our work with IETF leadership (IETF and IAB Chairs, the IETF
Executive Director)
and the RFC Editor to ensure that our access falls within their acceptable
use policies.

@@

\begin{figure}
  \centering
  \includegraphics[width=\figureWidthOneColumn]{figures/drafts-by-date.pdf}
  \caption{
    Number of active drafts over time. Internet-Drafts, the working draft
    series of the IETF, were introduced at the 13th IETF meeting in 1989.
  }
  \label{fig:drafts-by-date}
\end{figure}

@@

We observe that the number of active drafts correlates more strongly with
the number of active working groups (Figure \ref{fig:pub_wgs_yearly}) than
with RFC publication (Figure \ref{fig:rfcs_by_area}).  This is largely to
be expected: an Internet-Draft is a work-in-progress document that is
proposed for adoption by, and discussion in, an IETF working group prior to
eventual publication as an RFC.  Competing proposals are often submitted as
drafts, to allow a working group to review them and make a decision how to
proceed, and hence not all drafts become RFCs. As we discuss in
\S\ref{sec:success-factors:documents}, adoption is a significant
milestone.

@@

Figure~\ref{fig:keyword_usage_rates} further confirms the growing complexity
of RFCs, showing how the use of normative language has evolved.
Keywords are used in RFCs to indicate the normative requirements
an RFC imposes on implementations \cite{RFC2119}. Figure~\ref{fig:keyword_usage_rates}
shows the total number of occurrences of each of the 
keywords (i.e., MUST, MUST NOT, REQUIRED, SHALL, SHALL NOT, SHOULD, SHOULD
NOT, RECOMMENDED, MAY, OPTIONAL), divided by the page count of the RFC. As
shown, the median number of keywords per page grew from 2001 through to
2010, indicating a growing number of requirements being,
before plateauing in recent years.

@@

These shifts in affiliation reflect changes in fortune and commercial
success of the companies. They also show the ongoing importance of IETF
standards, with new companies choosing to send participants: while we do
not demonstrate any causal link, it is encouraging that commercially
successful companies opt to enable their employees to actively participate
in the IETF. Care is needed to ensure that this relevance is maintained,
however: the author pool has grown less diverse in terms of
companies that are represented. 35.4\% of authors came from the top 10
affiliations in the dataset in 2020, compared with 25.6\% in 2001.

@@

As we showed in Section \ref{sec:org-dyn:participants}, protocol
standardisation is an inherently social process with most day-to-day work
happening on public mailing lists.  We are specifically
interested in better understanding how influence is distributed across
stakeholders and how it might affect the standardisation process.  This is
of critical societal importance: the IETF has a major impact on global
Internet technologies, and understanding the social processes involved
would give us insight into not only the driving forces behind
standardisation, but also its resilience to the loss of major participants.
In this section we consider the following:
\begin{enumerate}
  \item How centralised is the active IETF community, and to what extent
    is it reliant on a small core of participants? 
  \item How do the most influential participants behave? 
  \item How does influence (determined by mailing list participation)
    relate to wider impacts throughout the IETF?
  \item Does the organisational affiliation of participants also
    influence the innovation (adoption of new work) within IETF?
\end{enumerate}

@@

We have identified that influential authors write more drafts
(\S\ref{subsec:impact_influence}), but are these drafts more likely to
see success?  That is, does the presence of a well-connected, influential,
author make it more likely that an Internet-draft is adopted by a working
group for review and development\footnote{There are occasional document,
known as Area Director sponsored drafts, that are published as standards
track RFCs without going through the working group process, but these too
rare ($\sim$3\% of RFCs) to significantly affect our analysis.} and
eventually published as a standards track RFC.  Using information from the
IETF Datatracker we compile a dataset for the \emph{adoption} of a draft by
a working group, comprising 11,632 drafts, and employ logistic regression to
analyse factors that affect working group adoption of draft.

@@

  \item We remove collinearity by using the Variance Inflation Criterion
    (VIF), removing all features with VIF $>5$.

@@

We further see that some of the topics extracted by LDA are also useful.
For example, Topic 13 is characterised by a cluster of terms associated
with MPLS, a widely deployed routing protocol. We find a negative
correlation with this topic, but this is likely because there is a
significant number of RFCs that propose modifications or additional
features for the protocol, some of which do not see deployment. Overall,
most of the results are in line with expectations, but more annotated data
is needed for further insights.

@@

\pb{Documents, Demographics, Communication Patterns:}
To the best of our knowledge, were the first to perform a comprehensive
statistical analysis of IETF activities considering trends in document
publication (\S\ref{sec:trends-documents}), author demographics
(\S\ref{sec:trends-demographics}), and communication patterns.
\S\ref{sec:org-dyn:participants}. That said, there have been several
prior efforts that have focused on specific aspects of the IETF.
Arkko, a former Chair of the IETF, maintains a website
(\url{https://www.arkko.com/tools/stats.html}) that provides various
statistics about the IETF, including about its documents, authors and their
affiliations. The Datatracker also provides some statistics on documents
and authors (\url{https://datatracker.ietf.org/stats/}).  Attendance
statistics are reported in the IETF plenaries (available in the proceedings
at \url{https://www.ietf.org/meeting/past/}).

@@

\pb{Influence and Impact:}
Our work to characterise participant influence and impact (\S\ref{sec:org-dyn:influence})
builds on a rich literature regarding influence in online social networks
and other communities including
studies of Twitter~\cite{cha2010measuring,weitzel2012measuring,anger2011measuring},
Instagram~\cite{zarei2020characterising} and decentralised social networks
\cite{harissigmetrics, hassanpleroma}. For example, \cite{ye2010measuring}
measured the impact of follower influence on message propagation. Others
have focused on devising metrics to capture influence: similarly
to~\cite{kourtellis2013identifying,weitzel2012measuring,sole2014centrality},
we use betweenness centrality as a metric of influence.

@@

\pb{Temporal Patterns in Communication:}
Analysing the temporal patterns in communication, as we do for the IETF
organisational hierarchy in \S\ref{sec:org-dyn:hierarchy}, is an emerging
field of research~\cite{masuda2016guide}.
Prior work has analysed the time evolution of important nodes in a
network~\cite{fire2020rise}, whether their neighbours are similarly
important~\cite{pedreschi2022temporal} and some others the correlations of
the trajectory of snapshots of networks over
time~\cite{lacasa2022correlations}. Compared to this, we brought new
analysis techniques, that expand on the idea of tracking the evolution of
node importance, to an email list corpus.

@@

The effects of hierarchy on the ``voice" of individuals within an
organisation is a key aspect~\cite{pfrombeck2023hierarchy} of
Organisational Behaviour (OB) research. A steep~\cite{hussain2019voice}
hierarchy can have a negative effect on voice by reducing the variety of
perspectives~\cite{anderson2010functions} in higher levels, or positive
effect where decisions can be made quickly. However, a diffused hierarchy
may also have negative effects~\cite{hussain2019voice} on voice by
contributing to lower level ``bystanders" who let the higher levels do the
talking, or a positive effect with a variety of approaches to management.
Similarly, a large ``power distance" can cause problems to lower level
voice if the higher levels are
``authoritarian"~\cite{duan2018authoritarian}, however if they are
``benevolent"~\cite{li2021does,guo2020inclusive} then power distance is
less of a problem. We tested these ideas in regard to the voluntary
and collaborative nature of the IETF organisation, whereas the literature
mostly focuses on traditional organisations.

@@

%..................................................................................................
\pb{Success Factors:}
The work in \S\ref{sec:success-factors} builds, in part, on Nikkhah et
al.~\cite{nikkhah2017statistical} by statistically exploring RFC adoption.
We extended this work with improved modelling, and by incorporating
additional features that characterise not only the RFCs themselves but also
the standardisation process that produced them.

@@

We explored the key standardisation activities within the IETF, one of the
primary technical standards developing organisations for the Internet,
considering trends in document production, complexity, and correctness
(\S\ref{sec:trends-documents}); the people involved in Internet standards
development and how the demographics of the community have shifted over
time (\S\ref{sec:trends-demographics}); interaction between participants,
who has influence and impact in the IETF and how that influence is evident
in their activities, and the dynamics of communications within the IETF
(\S\ref{sec:org-dyn}); use of language, to explore whether there are
differences in linguistic patterns and language use in more influential
participants, and to consider how the use of language reflects the
consensus-driven nature of the standards development process
(\S\ref{sec:language}); and factors that lead to success, exploring what
makes a document likely to be published as an RFC and what makes a
successful participant (\S\ref{sec:success-factors}).

@@

  \item While authorship is becoming more diverse as the IETF grows, there
    are still significant issues with representation from certain regions,
    including Africa and South America. If the IETF is to become more
    representative, then further effort is clearly required to build more diverse
    engagement, both in terms of geography, organisational affiliation, and
    gender diversity \cite{kaeo:2023:women}.
